---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
- 1. Load the data (i.e. read.csv())
```{r loading data}
activity_data <- read.csv('activity.csv')  
```

- 2. Process/transform the data (if necessary) into a format suitable for your analysis.
```{r format date}
activity_data$date <- as.Date(activity_data$date, "%Y-%m-%d")
```

## What is mean total number of steps taken per day?

- 1. Make a histogram of the total number of steps taken each day.

```{r histogram for total number of steps}
library(lattice)
steps_day <- aggregate(steps ~ date, data = activity_data, sum, na.rm = TRUE)
hist(steps_day$steps, main="", xlab = "Total number of steps per day", ylab="Frequency", col = "blue")

```

- 2. Calculate and report the mean and median total number of steps taken per day.

```{r mean and median calculations}
mean_steps_day <- as.integer(mean(steps_day$steps))
median_steps_day <- as.integer(median(steps_day$steps))
```
The mean number of steps taken per day is `r mean_steps_day`.

The median number of steps taken per day is `r median_steps_day`.

## What is the average daily activity pattern?

- 1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r time series plot across all days}
steps_interval <- aggregate(steps ~ interval, data = activity_data, mean, na.rm = TRUE)
plot(steps ~ interval, data = steps_interval, type = "l", xlab="5-minute interval",
     ylab="Average number of steps taken")
```

- 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r find interval with max number of steps}
max_interval <- steps_interval[which.max(steps_interval$steps), ]$interval
max_interval
```
The maximum number of steps is found in interval number `r max_interval`.

## Imputing missing values

- 1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r number of missing values}
missing_values <- sum(is.na(activity_data$steps))
missing_values
```
There are a total of `r missing_values` missing values.

- 2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

Step 1: Utilize the impute function in the Hmisc library.
Step 2: Copy the dataset to a new dataset name.
Step 3: Execute the impute function.
 
```{r impute missing data strategy}
require(Hmisc)
activity2 <- activity_data
activity2$steps <- impute(activity_data$steps, fun=mean)
```

- 3. Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r create new dataset with missing data filled}
require(Hmisc)
activity_data_new <- activity_data
activity_data_new$steps <- impute(activity_data$steps, fun=mean)

```
- 4.0 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

```{r histogram of for imputed dataset}
steps_day2 <- aggregate(steps ~ date, data = activity_data_new, sum, na.rm = TRUE)
hist(steps_day2$steps, main="", xlab = "Total number of steps per day", ylab="Frequency", col = "red")

```

```{r mean and median of imputed dataset}
mean_steps_day2 <- as.integer(mean(steps_day2$steps))
median_steps_day2 <- as.integer(median(steps_day2$steps))
```
The mean number of steps taken per day is `r mean_steps_day2`.

The median number of steps taken per day is `r median_steps_day2`.


- 4.1 Do these values differ from the estimates from the first part of the assignment? 

The mean value appears to be the same, whereas the median appears to be greater by 1 step in the second imputed dataset.  

- 4.2 What is the impact of imputing missing data on the estimates of the total daily number of steps?

The impact of imputing missing value largely depends on the approach of the analyst, but in this case the median increased slightly. 

## Are there differences in activity patterns between weekdays and weekends?

- 1. Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r createa new factor}
activity_data_new$day <- ifelse(as.POSIXlt(activity_data_new$date)$wday %in% 
                                   c(0,6), 'weekend', 'weekday')
```


- 2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the figure in the assignment instructions, which was created using simulated data:

```{r make a panle plot with two graphs}
steps_interval2 = aggregate(steps ~ interval + day, activity_data_new, mean) 
xyplot(steps ~ interval | factor(day), data = steps_interval2, type = "l",
       layout = c(1, 2), xlab = "5-minute interval", ylab = "Average number of steps taken") 

```